Computing Vowel Elision in Yoruba Collocations

### 2023-05-08 paper.tex
Spruce up conclusion a bit

The phenomenon of vowel hiatus in Yoruba, a language in the Niger-Congo family
spoken primarily in West Africa, is well-documented. A mature body of work
exists that explores the various aspects of it, largely from an
optimality-theoretic perspective \cite{ola2002yoruba, seidl2000yoruba,
pulleyblank1988vowel}. Two particularly well-studied instances of this problem
are the \textit{assimilation} case and the \textit{deletion} case. These are the
two primary strategies employed by the language for resolving sequences of two
or more vowels that occur at the boundary of two adjoined morphemes
(sequences of two vowels do not generally appear in Yoruba words due to
syllable structure constraints). This text will specifically explore the
second strategy of deletion, and in doing so, will specially attempt to provide
an analysis that is simple enough to be feasible to verify computationally. That
is, once we have developed our analysis, the core contribution of this work
will be to encode it as a computer program so it can be quickly and easily
verified whether it works for a range of corpus data. We employ an
optimality-theoretic perspective throughout, and provide a thorough discussion
in Section \ref{sec:computing-optimal-surface-forms} of the prior work on
computational interpretations of optimality theory. Afterwards, we describe
our own approach to automating our analysis and discuss its efficacy.

@@

\subsection{Analysis on corpus data}

@@

In this paper, we presented a classic Optimality-Theoretic analysis of
conjoined vowel elision in Yoruba. We then implemented it as a computer
program, finding a subtle issue in the analysis. With these issues addressed in
the implementation, our program outputs the correctly vowel-elided surface
form, given an underlying representation. Future work could go in several
interesting directions. Within the domain of Yoruba, the original analysis and
its implementation could be extended to handle the assimilatory strategy for
dealing with sequences of vowels at morpheme boundaries (as discussed in
\cite{ola2002yoruba}), as well as tone. This would likely involve extending
surface candidate generation to the full idealized algorithm described in
Section \ref{subsec:approach}, contending with epenthesis, metathesis, and
segment changes in addition to deletion. The implementation, as it exists,
faithfully models the parts of the algorithm in \ref{subsec:approach} that are
necessary in the context of Yoruba, so this would largely be a case of
extending the existing machinery.

>>>>

The phenomenon of vowel hiatus in Yoruba, a language in the Niger-Congo family
spoken primarily in West Africa, is well-documented. A mature body of work
exists that explores the various aspects of it, largely from an
optimality-theoretic perspective \cite{ola2002yoruba, seidl2000yoruba,
pulleyblank1988vowel}. Two particularly well-studied instances of this problem
are the \textit{assimilation} case and the \textit{deletion} case. These are the
two primary strategies employed by the language for resolving sequences of two
or more vowels that occur at the boundary of two adjoined morphemes
(sequences of two vowels do not generally appear in Yoruba words due to
syllable structure constraints). This text will specifically explore the
second strategy of deletion (in particular, ignoring tone markers as a related
but orthogonal case), and in doing so, will specially attempt to provide
an analysis that is simple enough to be feasible to verify computationally. That
is, once we have developed our analysis, the core contribution of this work
will be to encode it as a computer program so it can be quickly and easily
verified whether it works for a range of corpus data. We employ an
optimality-theoretic perspective throughout, and provide a thorough discussion
in Section \ref{sec:computing-optimal-surface-forms} of the prior work on
computational interpretations of optimality theory. Afterwards, we describe
our own approach to automating our analysis and discuss its efficacy.

@@

\subsection{Performance Analysis on Corpus Data}

@@

In this paper, we presented a classic Optimality-Theoretic analysis of
conjoined vowel elision in Yoruba. We then implemented it as a computer
program, finding a subtle issue in the analysis in doing so. With these issues
addressed in
the implementation, our program outputs the correctly vowel-elided surface
form, given an underlying representation. Future work could go in several
interesting directions. Within the domain of Yoruba, the original analysis and
its implementation could be extended to handle the assimilatory strategy for
dealing with sequences of vowels at morpheme boundaries (as discussed in
\cite{ola2002yoruba}), as well as tone. This would likely involve extending
surface candidate generation to the full idealized algorithm described in
Section \ref{subsec:approach}, contending with epenthesis, metathesis, and
segment changes in addition to deletion. The implementation, as it exists,
faithfully models the parts of the algorithm in \ref{subsec:approach} that are
necessary in the context of Yoruba, so this would largely be a case of
extending the existing machinery. To conclude, we have found that encoding our
conventional Optimality-Theoretic analysis computationally was highly effective
at identifying weaknesses and oversights, and at affirming its soundness after
addressing them. Optimality Theory lends itself quite naturally to computational
interpretations, and given our encouraging results, we look forward to the rich
body of work that lies ahead.
